Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-10-09 paper1_jiturra.Rmd
ajustes patrick 3

---
title: "Class-based network segregation, Economic Inequality and Redistributive Preferences across societies"
css: "input/css/custom.css" # custom CSS para html
fontsize: 12pt
linestretch: '1.5'          # interlineado  
#linestretch: '2.0'          # interlineado 
papersize: a4
link-citations: yes         # citas linkeadas
lang: en-US
# date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
date: "`r format(Sys.Date(), '%d %b %Y')`"
author:
- name: Julio Iturra-Sanhueza
  affiliation: Bremen International Graduate School of Social Sciences
  email: jiturra@uni-bremen.de
abstract: | 
  | Rising economic inequality has increased attention on the link between class-based networks and redistributive preferences in contemporary societies. Previous research has shown that class differences in redistributive preferences tend to narrow in more unequal societies, primarily due to higher redistributive preferences within the service class. However, little is known about how economic inequality affects the relationship between class-based networks and redistributive preferences. Moreover, income inequality can reinforce the role of social class in shaping social networks, highlighting how the lower class becomes increasingly segregated from other classes. Experiencing homogeneous class-based networks leads to segregated lifeworlds, which widen the gap between social classes and reduce awareness of the living conditions and worldviews of others. Empirically, I expect that homogeneous upper-class networks may reduce empathy and solidarity toward those in need, resulting in lower support for redistribution. Conversely, homogeneous working-class networks increase marginalization and strengthen the demand for redistribution. At the country level, I anticipate that in more unequal societies, the relationship between class-based network segregation and redistributive preferences weakens, due to the narrower differences between the upper and lower classes. This study examines the relationship between class-based network segregation and redistributive preferences and how it is moderated by economic inequality, utilizing cross-national data from 31,694 individuals across 31 societies. In line with these expectations, the findings suggest that social class shapes the association between network homogeneity and redistributive preferences: working-class homogeneity drives stronger redistributive preferences, while greater upper-class homogeneity reduces support for redistribution. Additionally, the conditional relationship between network segregation and social class weakens as inequality increases, particularly in the service class. The implications for studying class-based networks and political attitudes are also discussed.
  | 
  | **Keywords**: social networks, segregation, social class, income inequality, redistributive preferences 
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    keep_tex: false
    number_sections: true
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion  
  bookdown::html_document2:
    toc: false
    code_folding: hide
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false

always_allow_html: true      
linkcolor: gray                         # enlaces y citas en color azul
bibliography: input/bib/rgroup-redist.bib     # bibliografia en bibtex
csl: input/bib/apa-no-doi-no-issue.csl
editor_options:
  chunk_output_type: console            # en RStudio, mostrar output en consola
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm" # márgenes de página
header-includes:
  - \usepackage{times}   # Times New Roman
  # - \usepackage{helvet}
  - \usepackage{endnotes}
  - \let\footnote=\endnote 
  # - \renewcommand{\familydefault}{\sfdefault}
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \setlength\parindent{24pt} 
  - \usepackage{fancyhdr}
  - \fancyhead{} 
---

```{r render, include=FALSE, eval=FALSE}
rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::pdf_document2")
rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::html_document2")
rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::word_document2")
```

@@

**Word count**:  `r wordcount$words`

@@

Over the past few decades, cross-national studies on social class and redistributive preferences have predominantly centered on individual class positions [@lindh_class_2020]. Recently, however, a growing body of research has begun to focus on how the class profile of social networks influences redistributive preferences [@lindh_missing_2021; @lee_consider_2023; @paskov_crossclass_2022]. Although prior studies have shown that class-based network segregation—defined as the concentration of similar-class network ties—can undermine attachment to society [@otero_lives_2022], there is still limited understanding of how being embedded in a homogeneously class-based network impacts attitudes toward inequality and redistribution [@otero_power_2023]. Contributing to this debate, this research first aims to investigate the individual-level relationship between class-based network segregation and redistributive preferences. In addition, it has been suggested that social classes tend to form segregated networks according to the homophily principle, understood as the higher chances of forming social ties with others with similar characteristics [@mcpherson_birds_2001;@otero_lives_2022]. Furthermore, cross-national comparisons indicate that class-based network stratification strengthens in societies with high economic inequality, which in turn drives stronger social distance between classes, particularly among the lower classes [@otero_differences_2023; @pichler_social_2009]. Against this backdrop, the present study seeks to contribute to understanding how cross-national differences in income inequality act as a moderating factor in the relationship between class-based network segregation and redistributive preferences.

@@

Empirically, the class divide in redistributive preferences is well documented [@lindh_public_2015; @curtis_how_2015; @langsaether_more_2020; @brooks_why_2010]. Class-based explanations of redistributive preferences have predominantly focused on individuals. According to self-interest-driven theories, economic resources or exposure to risk explain why working classes with fewer resources and greater job insecurity tend to support redistribution more than the upper classes [@meltzer_rational_1981; @rehm_risks_2009]. Furthermore, while material interests often dominate in conditions of scarcity, value-based motivations, such as egalitarianism, may drive stronger support for redistribution under conditions of greater security and weaken under material hardship [@Kulin2013; @Maldonadoetal2019].

@@

Given this context, studying the connections between social classes from a network perspective offers a comprehensive way to understand redistributive preferences beyond the individual level. By focusing on network ties, the collective dimension of class relations is brought to the forefront, emphasizing that social ties encompass both social and economic resources [@lin_access_1986], which serve as the foundation for communal action [@weber_class_2011, pp. 57-59].

@@

Theoretically, class relations can be understood as the structure of social ties between different classes within the broader social system, represented by networks spanning various social strata [@blau_macrosociological_1977]. Research in social network literature consistently demonstrates that homophily—the tendency for individuals to associate with others who are similar—is a structured and persistent feature of social relations [@mcpherson_birds_2001]. For instance, friendships and family ties often display homogeneity in terms of social status or demographic characteristics, while more distant ties tend to connect individuals to different social groups, thereby contributing to network diversity [@diprete_segregation_2011; @lazarsfeld_friendship_1954]. Moreover, it is well established that socialization preferences play a role in forming segregated networks [@visser_attitudes_2004; @homans_human_1951]. However, this research aligns with the prevailing view that attitudinal similarity within networks arises primarily from structural contact opportunities shaped by the class composition of social ties, rather than from homophilic preferences for socializing with like-minded individuals [@feld_focused_1981].

To investigate the implications of networks for redistributive preferences, I propose distinguishing between two closely related—but distinct—perspectives on how class-based networks manifest within social structures.

First, the concept of _network diversity_ refers to the degree of connectedness to dissimilar ties (e.g., different occupations or activities), representing vertical access to diverse resources within social networks [@lin_access_1986]. In this context, diversity reflects the dispersion of different ties (alteri) across the network, independent of the individual’s (ego) characteristics [@otero_power_2023]. Empirical studies, such as those by @pichler_social_2009 and @lancee_income_2012, argue that higher civic engagement in formal organizations increases the likelihood of the upper classes forming connections with more diverse individuals, whereas the working classes tend to engage in more homogeneous participation. Similar patterns emerge in social tie composition, where the upper and intermediate classes maintain more diverse and prestigious social environments than the working class [@cepic_how_2020; @carrascosa_class_2023].

The second and primary perspective adopted in this study is _network segregation,_ defined as having contact primarily with people of the same class. This approach measures the similarity between an individual’s (ego) network ties and is conceptually closer to homophily, as it is rooted in an individual’s class position and has been empirically examined through network homogeneity [@otero_open_2021]. Evidence suggests that property-based boundaries are far less permeable than authority-based ones in the formation of cross-class ties. For example, @wright_relative_1992 suggests that class interests tend to widen the social distance between proprietors and manual workers, while the intermediate position of supervisors and their more frequent contact with manual workers make cross-class friendships more likely. Similarly, @otero_open_2021 identified a U-shaped pattern of acquaintance network segregation in Chile, where the middle classes (e.g., lower managerial professionals, clerks, and manual supervisors) exhibit lower network homogeneity. Likewise, segregation is most prevalent on both ends of the class structure, particularly among the lower classes. They argue that segregation in the lower classes is often driven by limited life chances and reduced social participation, while upper-class segregation is largely explained by self-selection—a practice that reinforces privileged positions, which can coexist with a broader range of social connections [@otero_open_2021].

@@

In addition to rational self-interest or value-based explanations, the argument that social ties have implications for attitude formation is not precisely new. More specifically, two approaches have discussed the role of social relations in redistributive preferences: reference groups and class-based networks.

@@

This research adopts the second approach, which suggests that social networks provide a comprehensive picture of the class relations that contribute to group identity formation and internalization of social norms [@kalmijn_social_2007]. Specifically, it has been argued that redistributive preferences are influenced not only by individuals' social class but also by the class positions of their network ties [@paskov_crossclass_2022]. Thus, opinions can either align or divide through social influence processes depending on the class positions of contacts and the level of network segregation [@lindh_missing_2021]. These arguments reflect the notion that classes are characterized as collectivities with varying degrees of cohesion and solidarity, comprising asymmetric status-based interactions related to material resources, cultural practices, and political preferences [@morris_attenuation_1996]. Following @sachweh_moral_2012, social integration can be impeded in societies with few opportunities for contact between different social classes, creating an "empathy gulf" that hinders individuals from understanding others' lifestyles amid rising inequality. Consequently, segregated interactions may lead individuals to perceive the lives of different classes as more distant and undermine feelings of social integration [@sachweh_moral_2012]. Thus, segregation can potentially diminish social cohesion as it exacerbates perceptions of others as strangers which in turn reduces empathy and solidarity [@otero_lives_2022].

How do network ties affect redistributive preferences? The class position of surrounding family members, friends, and acquaintances affects support for redistribution, as these ties function as socialization agents whose impact can be amplified in segregated social networks. In principle, political attitudes are connected to class interests and norms as they are socialized in the family of origin during childhood and early adulthood. For instance, @lee_consider_2023 shows that individuals with network ties to the upper class through parental connections tend to support redistribution and progressive taxation less than those from working-class family backgrounds. Moreover, since households share risk based on the class position of their members, redistributive preferences are shaped not only by family background but also by the class positions of partners. For example, @paskov_crossclass_2022 indicates that working-class ties bolster redistributive preferences, whereas ties with the upper class decrease them, with the effects becoming more pronounced when the class positions of individuals, partners, and parents form a more homogeneous network. Similarly, @lindh_missing_2021 found that friendship and acquaintanceship ties to the managerial class are associated with lower redistributive preferences compared to ties with the sociocultural and working classes. Hence, this suggests that individuals tend to adjust their attitudes based on the class position of their contacts [@lindh_missing_2021].

In summary, I anticipate a weak direct association between network segregation and redistributive preferences. This is mainly because network homogeneity - as the proportion of similar class network ties - does not distinguish between anchor class positions and refers to the overall degree of segregation. Conversely, I hypothesize that the association of network homogeneity with redistributive preferences is conditional on social class as homogeneous social networks should reinforce attitude similarity (*segregation hypothesis*). Specifically, I propose that greater network segregation in the lower classes is associated with higher redistributive preferences, whereas greater segregation in the upper classes is related to lower redistributive preferences. Therefore, the first hypothesis is as follows:

@@

The literature on the role of economic inequality on (1) the class-attitude link and (2) social relations have been independently developed.

Theoretically, studies on economic inequality, social class, and redistributive preferences show varying class reactions to rising inequality. Political economists suggest that high-income individuals are far from being monolithic in their redistributive preferences, motivated by concerns about inequality's adverse effects (e.g., crime), leading to altruistic support for redistribution [@dimick_models_2018; @dimick_altruistic_2017; @rueda_externalities_2016]. Conversely, the moral economy literature in sociology explains affluent differences through evaluations of distributive justice regarding resource allocation procedures [@liebig_sociology_2016]. At the same time, the literature on the role of income inequality on social cohesion has argued that unequal societies strengthen barriers to trustworthy social connections. Neckerman and Torche [-@neckerman_inequality_2007, p. 344] noted that marginalization is linked to reduced participation in social life, exacerbated by greater inequality. Also, rising economic inequality erodes social relations by reducing social trust, which underpins mutual understanding and collective solidarity [@kragten_income_2017]. Conversely, it is noted that societies with low inequality and comprehensive welfare states enhance civic and social participation, increasing generalized trust and reducing social conflict [@uslaner_inequality_2005]. Similarly, it has been argued that social participation can nurture higher chances for cross-class interactions and motivate more egalitarian attitudes, being upper-class citizens who support stronger collective solidarity through government redistribution in more cohesive communities [@yamamura_social_2012].

Empirically, it has been shown that affluent groups are more responsive, perceiving that rising inequality affects the overall opportunity structure and social mobility chances [@sachweh_why_2019]. Also, higher perceived inequality of opportunity among the upper classes can motivate support for redistribution as a matter of procedural justice [@kim_socioeconomic_2018]. In contrast, low-income individuals view ascribed characteristics as more crucial in constraining opportunities, irrespective of current income inequality [@sachweh_why_2019]. Similarly, @edlund_democratic_2015 argue that in unequal societies with residual welfare states, social class lacks significant political meaning, weakening distributive conflicts around traditional working-class organizations (e.g., unions). Thus, stronger support for state-organized redistribution among the upper classes can be attributed to their greater awareness of the incentives and societal consequences of income inequality [@curtis_how_2015; @Svallfors2006]. At the same time, cross-national comparative evidence indicates that stratified access to social activities and networks is reinforced in unequal societies. For example, @pichler_social_2009 demonstrated that class differences in civic and family network participation are exacerbated in unequal societies. Similarly, @lancee_income_2012 found that income-based stratification in civic participation increases with inequality. Also, @otero_differences_2023 found that economic inequality enhances 
interdependence among cultural, economic, and social capital, which results in stronger stratified chances for cross-class interactions. Thus, in unequal societies, the upper classes can navigate diverse social settings while maintaining their distinct positions, whereas the lower classes remain marginalized and segregated due to others' choices [@otero_open_2021, p.24].

Against this backdrop, there are at least three emerging theoretical implications on the role of income inequality on class-based networks and redistributive preferences. The first implication is that it is plausible to assume that income inequality drives greater segregation, as the result of lower participation and social trust. However,  it is worth noticing that resource interdependency seems to exacerbate social exclusion for the lower classes, while the upper classes may hold more cohesive experiences that drive stable or even lower segregation in more unequal societies. A second implication is that the relationship between class and attitudes weakens in highly unequal societies, where the upper classes tend to show more support for redistribution, contrasting with the relatively stable preferences of the lower classes. In addition, it is also true that in egalitarian societies with more encompassing welfare states class-based distributive struggles tend to be more institutionalized (e.g. through unions or political parties). As a result, class positions are more closely implied in political conflict as the institutional framework allows them to better articulate their interests. A third implication is that as inequality increases and the class divide in redistributive preferences narrows, this may also be extended to the influence of class-based segregation on redistributive preferences. As the class-attitude link becomes mild in more unequal contexts, it is also expected that homogeneous class-based networks might not have the same strength where the role of class positions loses significance. Therefore, I expect income inequality to weaken the interaction between network homogeneity and social class. Specifically, I hypothesize that the conditional relationship between network homogeneity and social class stated at the individual level (*segregation hypothesis*) will be less pronounced in the context of higher income inequality (*mitigation hypothesis*). Given these considerations, the second hypothesis is as follows:

@@

I use two indicators to measure redistributive preferences. The first indicator corresponds to support for government redistribution and it is measured by the item: "It is the responsibility of the government to reduce the differences in income between people with high incomes and those with low incomes." The second indicator is egalitarian preferences and it is measured by the item: "For a society to be fair, differences in people's standard of living should be small." Both indicators use a five-point Likert scale with the following categories: 'Strongly agree' (1), 'Agree' (2), 'Neither agree nor disagree'(3), 'Disagree' (4), and 'Strongly disagree'(5) (*r* = 0.62). Following @svallfors_government_2013, I reverse-coded, averaged and normalized the indicators in a 0 to 100 index, where higher values reflect stronger redistributive preferences.

@@

To measure social class, I employ the Erikson-Goldthorpe-Portocarrero (EGP) class scheme [@Erikson1992]. The EGP scheme is one of the most consistent and validated measures for social class positions in comparative research and has demonstrated its validity in both industrialized and late-industrialized societies [@evans_political_2013;@barozet_measurement_2021]. Therefore, information about occupations, self-employment status, and the number of employees is used to classify respondents into six class positions. Following previous research, a simplified version of the EGP class scheme that collapses three classes is employed [@sosnaud_class_2013;@edlund_influence_2003]. Specifically, this version distinguishes among the Service Class (higher and lower managerial and professionals), Intermediate Class (routine nonmanual workers and self-employed), and Working Class (manual supervisors, skilled and unskilled manual workers) [^egpclass] (see Table \@ref(tab:tab-an-classag) in the Appendix).

[^egpclass]: Self-employed farmers (IVc) are grouped in the self-employed class, while agricultural labor (VIIb) is in the working class.

I employed the position generator as the basis for the class-based network homogeneity measure. This instrument has been widely used in social capital studies and follows an ego-centered approach where is assumed that social ties to different hierarchical positions in the social structure provide access to social resources [@vandergaag_position_2008]. The position generator employed included a list of ten occupations. Here, occupations are presented to the respondent in a grid that allows them to declare whether they know (or not) a person who performs that occupation. The tie can be clasified in four groups, defining tie to this person is as a "Family or relative," "Close friend,” "Someone else I know," or "No one.”  With this information, the first three categories are coded as 1 to represent the presence of a tie ("Knows") and 0 as the abscence of a conection to a person with that occupation ("Does not know").  Subsequently, all declared ties were summed to represent the total number of occupations known by the respondent.

>>>>

---
# title: "Class-based network segregation, Economic Inequality and Redistributive Preferences across societies"
# css: "input/css/custom.css" # custom CSS para html
fontsize: 12pt
linestretch: '1.5'          # interlineado
# #linestretch: '2.0'          # interlineado 
papersize: a4
link-citations: yes         # citas linkeadas
lang: en-US
# # date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
# date: "`r format(Sys.Date(), '%d %b %Y')`"
# author:
# - name: Julio Iturra-Sanhueza
#   affiliation: Bremen International Graduate School of Social Sciences
#   email: jiturra@uni-bremen.de
# abstract: | 
#   | Rising economic inequality has increased attention on the link between class-based networks and redistributive preferences in contemporary societies. Previous research has shown that class differences in redistributive preferences tend to narrow in more unequal societies, primarily due to higher redistributive preferences within the service class. However, little is known about how economic inequality affects the relationship between class-based networks and redistributive preferences. Moreover, income inequality can reinforce the role of social class in shaping social networks, highlighting how the lower class becomes increasingly segregated from other classes. Experiencing homogeneous class-based networks leads to segregated lifeworlds, which widen the gap between social classes and reduce awareness of the living conditions and worldviews of others. Empirically, I expect that homogeneous upper-class networks may reduce empathy and solidarity toward those in need, resulting in lower support for redistribution. Conversely, homogeneous working-class networks increase marginalization and strengthen the demand for redistribution. At the country level, I anticipate that in more unequal societies, the relationship between class-based network segregation and redistributive preferences weakens, due to the narrower differences between the upper and lower classes. This study examines the relationship between class-based network segregation and redistributive preferences and how it is moderated by economic inequality, utilizing cross-national data from 31,694 individuals across 31 societies. In line with these expectations, the findings suggest that social class shapes the association between network homogeneity and redistributive preferences: working-class homogeneity drives stronger redistributive preferences, while greater upper-class homogeneity reduces support for redistribution. Additionally, the conditional relationship between network segregation and social class weakens as inequality increases, particularly in the service class. The implications for studying class-based networks and political attitudes are also discussed.
#   | 
#   | **Keywords**: social networks, segregation, social class, income inequality, redistributive preferences 
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    keep_tex: false
    number_sections: true
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion  
  bookdown::html_document2:
    toc: false
    code_folding: hide
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false
  bookdown::word_document2:
    reference_docx: "input/mytemplate.docx"
    toc: false
    code_folding: hide
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false

always_allow_html: true      
linkcolor: gray                         # enlaces y citas en color azul
bibliography: input/bib/rgroup-redist.bib     # bibliografia en bibtex
csl: input/bib/apa-no-doi-no-issue.csl
editor_options:
  chunk_output_type: console            # en RStudio, mostrar output en consola
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm" # márgenes de página
header-includes:
  - \usepackage{times}   # Times New Roman
  # - \usepackage{helvet}
  - \usepackage{endnotes}
  - \let\footnote=\endnote 
  # - \renewcommand{\familydefault}{\sfdefault}
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage{graphicx}
  - \usepackage{setspace}  
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \setlength\parindent{24pt} 
  - \usepackage{fancyhdr}
  - \fancyhead{} 
---

```{r render, include=FALSE, eval=FALSE}
rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::pdf_document2")
rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::word_document2")
rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::html_document2")
```

@@

<!-- **Word count**:  `r wordcount$words` -->

@@

Over the past few decades, cross-national studies on social class and redistributive preferences have predominantly centered on individual class positions [@lindh_class_2020]. Recently, however, a growing body of research has begun to focus on how the class profile of social networks influences redistributive preferences [@lindh_missing_2021; @lee_consider_2023; @paskov_crossclass_2022]. Additionally, social classes tend to form segregated networks according to the homophily principle, understood as the higher chances of forming social ties with others with similar characteristics [@mcpherson_birds_2001;@otero_open_2021]. Although prior evidence has shown that class-based network segregation —as the proportion of network contacts from the same class relative to an individual's class position— can undermine attachment to society in highly unequal societies like Chile [@otero_lives_2022], there is still limited understanding of how being embedded in a homogeneously class-based network impacts attitudes toward inequality and redistribution in cross-national perspective [@otero_power_2023]. Contributing to this debate, this research first aims to investigate the extent individual-level relationship between class-based network segregation and redistributive preferences. Furthermore, cross-national comparisons indicate that class-based network stratification strengthens in societies with high economic inequality, which in turn drives wider social distance between classes [@otero_differences_2023; @pichler_social_2009]. Against this backdrop, the present study seeks to contribute to understanding how cross-national differences in income inequality act as a moderating factor in the relationship between class-based network segregation and redistributive preferences.

@@

Empirically, the class divide in redistributive preferences is well documented [@lindh_public_2015; @curtis_how_2015; @langsaether_more_2020; @brooks_why_2010]. Class-based explanations of redistributive preferences have predominantly focused on individuals. According to self-interest-driven theories, economic resources or risk exposure explain why working classes with fewer resources and greater job insecurity tend to support redistribution more than the upper classes [@meltzer_rational_1981; @rehm_risks_2009]. Furthermore, while material interests often dominate in conditions of scarcity, value-based motivations, such as egalitarianism, may drive stronger support for redistribution under conditions of greater security and weaken under material hardship [@Kulin2013; @Maldonadoetal2019].

@@

Given this context, studying the connections between social classes from a network perspective offers a comprehensive way to understand redistributive preferences beyond the individual level. By focusing on network ties, the collective dimension of between-class relations is brought to the forefront, emphasizing that social ties encompass both social and economic resources [@lin_access_1986], which serve as the foundation for communal action [@weber_class_2011, pp. 57-59].

@@

Theoretically, class relations can be understood as the structure of social ties between different classes within the broader social system, represented by networks spanning various social strata [@blau_macrosociological_1977]. Research in social network literature consistently demonstrates that homophily—the tendency for individuals to associate with others who are similar—is a structured and persistent feature of social relations [@mcpherson_birds_2001]. For instance, friendships and family ties often display homogeneity in terms of social status or demographic characteristics, while more distant ties tend to connect individuals to different social groups, thereby contributing to network diversity [@diprete_segregation_2011; @lazarsfeld_friendship_1954]. Moreover, it is well established that socialization preferences play a role in forming segregated networks [@visser_attitudes_2004; @homans_human_1951]. However, this research aligns with the prevailing view that attitudinal similarity within networks arises primarily from structural contact opportunities shaped by the class composition of social ties, rather than from homophilic preferences for socializing with like-minded individuals [@feld_focused_1981]. Thus, to investigate the implications of networks for redistributive preferences, I propose distinguishing between two closely related—but distinct—perspectives on how class-based networks are structured.

First, _network diversity_ refers to the degree of connectedness to dissimilar ties (e.g., different occupations or activities), representing access to diverse resources within social networks [@lin_access_1986]. Thus, refers to the dispersion of an attribute tie (alteri) within the network, independently of the characteristic that an individual has (ego) [@otero_power_2023]. Related studies such as @pichler_social_2009 and @lancee_income_2012, have argued that variety participation in formal organizations increases the chances of upper classes forming more diverse social connections. In contrast, the working classes show a more homogeneous participation repertoire. Similar patterns emerge in network socioeconomic composition, where the upper middle classes maintain more prestigious and diverse networks than the working classes [@cepic_how_2020; @carrascosa_class_2023]. However, I argue that while diversity accounts for the dispersion of an attribute within the network this approach lacks an anchor position that allows to address segregation more precisely.

By contrast, the perspective adopted in this study is _network segregation,_ defined as the pattern of contact primarily with people of the same characteristics. The focus is on the similarity between the class position of individuals (ego) and their network ties (alteri). Therefore, it is conceptually closer to homophily and has been empirically examined through network homogeneity [@otero_open_2021]. Some studies have suggested that property-based boundaries are far less permeable than authority-based ones in the formation of cross-class ties. For example, @wright_relative_1992 suggests that class interests tend to widen the social distance between proprietors and manual workers, while the intermediate position of supervisors and their more frequent contact with manual workers make cross-class friendships more likely. Similarly, @otero_open_2021 identified a U-shaped pattern of acquaintance network segregation in Chile, where the middle classes (e.g., lower managerial professionals, clerks, and manual supervisors) exhibit lower network homogeneity. Likewise, segregation is most prevalent on both ends of the class structure, particularly among the lower classes. They argue that segregation in the lower classes is often driven by limited life chances and reduced social participation, while upper-class segregation is largely explained by self-selection—a practice that reinforces privileged positions, which can coexist with a broader range of social connections [@otero_open_2021].

@@

In addition to individual class-based explanations, the argument that social ties have implications for attitude formation is not precisely new. More specifically, two approaches have discussed the role of social relations in redistributive preferences: reference groups and class-based networks.

@@

This research adopts the second approach, which suggests that social networks provide a comprehensive picture of the class relations that contribute to group identity formation and internalization of social norms [@kalmijn_social_2007]. Specifically, it has been argued that redistributive preferences are influenced not only by individuals' social class but also by the class positions of their network ties [@paskov_crossclass_2022]. Thus, opinions can either align or divide through social influence processes depending on the class positions of contacts and the level of network segregation [@lindh_missing_2021]. These arguments reflect the notion that classes are characterized as collectivities with varying degrees of cohesion and solidarity, comprising asymmetric status-based interactions related to material resources, cultural practices, and political preferences [@morris_attenuation_1996]. Following @sachweh_moral_2012, social integration can be impeded in societies with few opportunities for contact between different social classes, creating an "empathy gulf" that hinders individuals from understanding others' lifestyles amid rising inequality. Consequently, segregated interactions may lead individuals to perceive the lives of different classes as more distant and undermine feelings of social integration [@sachweh_moral_2012]. Thus, segregation can potentially diminish social cohesion as it exacerbates perceptions of others as strangers which in turn reduces empathy and solidarity [@otero_lives_2022]. Although both approaches share the theoretical assumption that social networks contribute information (_availability heuristic_) and affect attitudes (_social influence_). I argue that an advantage of the network perspective is that it has been more effective in defining the structural position of network ties, and has more directly addressed the relationship with redistributive preferences.

How do network ties affect redistributive preferences? The class position of surrounding family members, friends, and acquaintances affects support for redistribution, as these ties are a source of social influence whose impact can be amplified in segregated social networks. In principle, political attitudes are connected to class interests and norms as they are primarily -  but exclusively - socialized in the family of origin during childhood and early adulthood. For instance, @lee_consider_2023 shows that individuals with network ties to the upper class through parental connections tend to support redistribution and progressive taxation less than those from working-class family backgrounds. Moreover, since households share risk based on the class position of their members, redistributive preferences are shaped not only by family background but also by the class positions of partners. For example, @paskov_crossclass_2022 indicates that working-class ties bolster redistributive preferences, whereas ties with the upper class decrease them, with the effects becoming more pronounced when the class positions of individuals, partners, and parents form a more homogeneous network. Beyond family ties, @lindh_missing_2021 found that friendship and acquaintanceship ties to the managerial class are associated with lower redistributive preferences compared to ties with the sociocultural and working classes. Hence, this suggests that individuals tend to adjust their attitudes based on the class position of their contacts [@lindh_missing_2021].

In summary, I expect a weak direct association between network segregation and redistributive preferences. This is mainly because network homogeneity - as the proportion of similar class network ties - does not distinguish between anchor class positions and refers to the overall degree of segregation. Conversely, I hypothesize that the association of network homogeneity with redistributive preferences is conditional on social class as homogeneous social networks should reinforce attitude similarity (*segregation hypothesis*). Specifically, I propose that greater network segregation in the lower classes is associated with higher redistributive preferences, whereas greater segregation in the upper classes is related to lower redistributive preferences. Therefore, the first hypothesis is as follows:

@@

Theoretically, rising income inequality prompts diverse responses across social classes regarding redistributive preferences and social cohesion. Political economy perspectives suggest that affluent individuals are not uniform in their attitudes, with some supporting redistribution out of concern for the broader societal effects of inequality, such as crime and instability [@dimick_models_2018; @dimick_altruistic_2017]. The moral economy approach in sociology explains this variation through differing views on distributive justice and fairness in resource allocation [@liebig_sociology_2016]. At the same time, research on social cohesion shows that inequality erodes trust and participation, particularly among marginalized groups, exacerbating social exclusion [@neckerman_inequality_2007; @kragten_income_2017]. In contrast, egalitarian societies with strong welfare states foster higher levels of civic engagement and cross-class interactions, which strengthen solidarity and promote more egalitarian attitudes [@uslaner_inequality_2005; @yamamura_social_2012].

Empirical evidence further indicates that affluent individuals are more likely to perceive inequality as a threat to social mobility and opportunity structures, motivating support for redistribution as a form of procedural justice [@sachweh_why_2019; @kim_socioeconomic_2018]. By contrast, low-income individuals tend to focus on structural barriers that limit opportunities, regardless of inequality [@sachweh_why_2019]. In highly unequal societies with residual welfare states, where traditional class-based political organizations like unions have weakened, upper-class support for redistribution stems from a heightened awareness of inequality’s societal consequences [@curtis_how_2015; @Svallfors2006; @edlund_democratic_2015]. Moreover, income inequality reinforces stratification in social participation, as the affluent maintain access to diverse networks while the lower classes face growing exclusion [@pichler_social_2009; @lancee_income_2012]. This stratification limits cross-class interactions and further entrenches social divisions, with the upper classes navigating diverse social environments while the lower classes remain segregated and marginalized [@otero_differences_2023; @otero_open_2021].

Against this backdrop, three key implications emerge regarding the role of income inequality in shaping class-based networks and redistributive preferences. First, income inequality likely leads to greater segregation due to reduced participation and lower social trust; however, while resource interdependency exacerbates social exclusion for lower classes, upper classes may experience greater cohesion, potentially stabilizing or even reducing segregation in more unequal societies. Second, the relationship between class and attitudes weakens in highly unequal contexts, where upper classes tend to show increased support for redistribution, contrasting with the relatively stable preferences of lower classes. In egalitarian societies with comprehensive welfare states, class-based distributive struggles become more institutionalized, allowing class positions to be closely tied to political conflict. Lastly, as inequality rises and the class divide in redistributive preferences narrows, the influence of class-based segregation on these preferences may diminish, with homogeneous class networks losing significance as class positions become less relevant in shaping attitudes.

Therefore, I expect income inequality to weaken the interaction between network homogeneity and social class. Specifically, I hypothesize that the conditional relationship between network homogeneity and social class at the individual level will be less pronounced in the context of higher income inequality (_mitigation hypothesis_). Given these considerations, the second hypothesis is as follows:

@@

I use two indicators to measure redistributive preferences. The first indicator is support for government redistribution measured by the item: "It is the responsibility of the government to reduce the differences in income between people with high incomes and those with low incomes." The second indicator is egalitarian preferences measured by the item: "For a society to be fair, differences in people's standard of living should be small." Both indicators use a five-point Likert scale with the following categories: 'Strongly agree' (1), 'Agree' (2), 'Neither agree nor disagree'(3), 'Disagree' (4), and 'Strongly disagree'(5) (*r* = 0.62). Following @svallfors_government_2013, I reverse-coded, averaged, and normalized the indicators in a 0 to 100 index, where higher values reflect stronger redistributive preferences.

@@

I employ the Erikson-Goldthorpe-Portocarrero (EGP) class scheme [@Erikson1992] to measure social class. The EGP scheme is the most consistent and validated measure for class positions in comparative research and has demonstrated its validity in both industrialized and late-industrialized societies [@evans_political_2013;@barozet_measurement_2021]. Therefore, information about occupations, self-employment status, and the number of employees is used to classify respondents into six class positions. Following previous research, a simplified version of the EGP class scheme that collapses three classes is employed [@sosnaud_class_2013;@edlund_influence_2003]. Specifically, this version distinguishes among the Service Class (higher and lower managerial and professionals), Intermediate Class (routine nonmanual workers and self-employed), and Working Class (manual supervisors, skilled and unskilled manual workers) [^egpclass] (see Table \@ref(tab:tab-an-classag) in the Appendix).

[^egpclass]: Self-employed farmers (IVc) are included in the self-employed class, while agricultural labor (VIIb) is in the working class.

I employed the position generator as the basis for the class-based network homogeneity measure. This instrument has been widely used in social capital studies and follows an ego-centered approach where is assumed that social ties to different hierarchical positions in the social structure provide access to social resources [@vandergaag_position_2008]. The position generator employed included a list of ten occupations. Here, occupations are presented to the respondent in a grid that allows them to declare whether they know (or not) a person who performs that occupation. The tie can be classified into four groups, defining tie to this person as a "Family or relative," "Close friend,” "Someone else I know," or "No one.”  With this information, the first three categories are coded as 1 to represent the presence of a tie ("Knows") and 0 as the absence of a connection to a person with that occupation ("Does not know").  Subsequently, all declared ties were summed to represent the total number of occupations known by the respondent.
